---
title: 'Pinniped Stranding Analysis Research Questions'
author: Amanda Warlick, March 2017
output:
  word_document:
    reference_docx: mytemplate.docx

---

```{r, include = F}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
```
 

```{r, include = F, echo = F}
library(ggplot2)
library(gplots)
library(tidyr)
library(ggmap)
library(data.table)
library(cowplot)
library(knitr)
library(reshape2)
library(stringr)
library(magrittr)
library(scales)
library(stats) 
library(PMCMR)
library(sciplot)
library(pgirmess) #kruskalmc()
library(MASS) #glm.nb()
library(multcomp) #glht()
library(dplyr)
detach("package:Hmisc")
# detach("package:multcomp")
# detach("package:TH.data")
detach("package:MASS")

setwd("~/Documents/R/Strandings")

# Data Load 
###############################################################################
# Data are pre-cleaned
pinnipeds_data <- read.csv("pinnipeds_data_protected.csv", header = TRUE, na.strings = "", stringsAsFactors = FALSE) %>%
  transform(Decade = ifelse(Year.of.Observation < 2000, '1990s', 
                            ifelse(Year.of.Observation < 2010, '2000s', '2010s'))) %>%
  transform(Period = ifelse(Year.of.Observation < 2005, 'Pre-Prescott', 'Post-Prescott')) %>%
  filter(Year.of.Observation > 1990)

# Set Up Factors
pinnipeds_data$Age.Class <- factor(pinnipeds_data$Age.Class, 
                                   levels = c("Pup", "Yearling", "Subadult", "Adult", "Unid"))

pinnipeds_data$Month.of.Observation <- factor(pinnipeds_data$Month.of.Observation, 
                      levels = c('JAN', 'FEB', 'MAR', 'APR', 'MAY', 'JUN', 'JUL', 'AUG', 'SEP', 'OCT', 'NOV', 'DEC'))

```


##Changes in total annual strandings per year  
 
- Are total annual combined strandings or of each species increasing or decreasing over time?
```{r}
### Are total annual combined strandings or of each species increasing over time?  

#Could use lm() to derive and report slope of line, glm for test statistics/p-values, or glm and reporting fold increase per decade.

## (a) All strandings, combined species
data_by_year <- pinnipeds_data %>% 
  group_by(Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

sctest_data <- data_by_year %>%
  arrange(cnt)

sctest(Year.of.Observation ~ cnt, type = "Chow", point = 12, data = sctest_data)

data_by_decade <- pinnipeds_data %>% 
  group_by(Decade) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# GLM with Poisson
summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = data_by_year))
#slope of line =  exp(0.0446*x) = y = 1.045^x = 4.5%/year

#Negative binomial GLM
summary(glm.nb(cnt ~ Year.of.Observation, data = data_by_year))
#slope of line =  exp(0.046800*x) = y = 1.0479^x = 4.8%/year
#new: exp(0.044076) = y = 1.045062^x = 4.5%/year

#Linear
# summary(lm(cnt ~ Year.of.Observation, data_by_year)) #y = 25x


## (b) All strandings, separate species
all_sp <- pinnipeds_data %>%
  group_by(Year.of.Observation, Pinniped.Common.Name) %>%
  summarize(cnt = n_distinct(National.Database.Number))

###Harbor seal
#Linear
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Harbor seal'))) #y = 16x

#GLM with Poisson
summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'Harbor seal'))) # y = 1.1^x

#Negative binomial - use to avoid concerns of overdispersion
summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Harbor seal'))) # y = exp(0.04428) = 1.04527 = 4.5% per year

###CSL
#Linear
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'California sea lion'))) # y = 7.9x

#GLM with Poisson
# summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'California sea lion'))) # y = 1.1^x

#Negative binomial
summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'California sea lion'))) # exp(0.0744) = y = 1.077^x 

###SSL
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) # y = 3.9x

# summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) # y = 1.1^x

summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) # y = exp(0.11959x) = 1.127^x = 12%

#GFS
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) # y = 1.4x

# summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) # y = 1.1^x

summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) # y = exp(0.1159x) = y = 1.122^x = 12%

###NES
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Northern elephant seal'))) 

summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'Northern elephant seal')))

summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Northern elephant seal'))) #not sig

###NFS
# summary(lm(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Northern fur seal'))) 

summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = all_sp %>% filter(Pinniped.Common.Name == 'Northern fur seal')))

summary(glm.nb(cnt ~ Year.of.Observation, data = all_sp %>% filter(Pinniped.Common.Name == 'Northern fur seal'))) #exp(0.0473) = y = 1.0484

```

- Are total annual *number* of HI cases increasing over time?
```{r}
# Are total annual human interaction cases increasing/decreasing over time for:

# (a) combined species 
HI_by_year <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# <--- Yes, significant increase of all HI cases, 
# summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = HI_by_year))
#regression equation: y = exp(0.1089*x) = 1.11^x

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_by_year))
#regression equation: y = exp(0.07257) = 1.07^x = 7% per year

# (b) certain HI types
HI_type_by_year <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Year.of.Observation, Interaction.Type) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = HI_type_by_year %>% filter(Interaction.Type == 'Gunshot')))
summary(glm.nb(cnt ~ Year.of.Observation, data = HI_type_by_year %>% filter(Interaction.Type == 'Gunshot'))) #exp(0.04955) = y = 1.051

summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = HI_type_by_year %>% filter(Interaction.Type == 'Fisheries')))
summary(glm.nb(cnt ~ Year.of.Observation, data = HI_type_by_year %>% filter(Interaction.Type == 'Fisheries')))  #exp(0.08605) = y = 1.068

summary(glm(cnt ~ Year.of.Observation, family = poisson(link = log), data = HI_type_by_year %>% filter(Interaction.Type == 'Boat')))
summary(glm.nb(cnt ~ Year.of.Observation, data = HI_type_by_year %>% filter(Interaction.Type == 'Boat'))) #exp(0.07688) = 1.0799

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_type_by_year %>% filter(Interaction.Type == 'Other'))) #exp(0.135) y = 1.14x

# (c) for individual species?  
HI_sp <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Year.of.Observation, Pinniped.Common.Name) %>%
  summarize(cnt = n_distinct(National.Database.Number))

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'Harbor seal'))) # y = 1.073^x

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'California sea lion'))) # y = 1.075^x

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) # y = 1.096^x

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'Northern elephant seal'))) # not sig

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) # y = 1.110^x

summary(glm.nb(cnt ~ Year.of.Observation, data = HI_sp %>% filter(Pinniped.Common.Name == 'Northern fur seal')))  #insufficient observations?
```

- Is the *prevalence* of human interaction cases increasing or decreasing over time for certain species? 
```{r}

#Changing prevalence over time - by year and by period

# (a) All HI cases, species combined

###YEAR####

HI_yes_no <- pinnipeds_data %>%
  group_by(Year.of.Observation, Findings.of.Human.Interaction) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  dcast(Year.of.Observation ~ Findings.of.Human.Interaction, value.var = 'cnt')
HI_yes_no[is.na(HI_yes_no)] <- 0
HI_yes_no <- HI_yes_no %>%
    transform(Yes = Y, No = (N+CBD)) %>%
  dplyr::select(-c(CBD, N, Y)) 

#link = logit default
summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial(link = logit), data = HI_yes_no)) # y = 0.046x = 0.05 per year untransformed - prop change over time

#summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial(link = log), 
#             data = HI_yes_no)) # y = 1.059^x = 5.9%/year 


####PERIOD####
HI_yes_no_period <- HI_yes_no %>%
  transform(period = ifelse(Year.of.Observation < 2000, '1991-1999',
                            ifelse(Year.of.Observation < 2010 & Year.of.Observation > 1999, '2000-2009', '2009-2016'))) %>%
  group_by(period) %>%
  summarize(Yes = sum(Yes), No = sum(No))

summary(glm(cbind(Yes, No) ~ period, family = binomial(link = logit), data = HI_yes_no_period)) 


# (b) All HI cases, separate species
HI_yes_no_sp <- pinnipeds_data %>%
  filter(Pinniped.Common.Name != 'Unidentified') %>% 
  group_by(Pinniped.Common.Name, Year.of.Observation, Findings.of.Human.Interaction) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  dcast(Pinniped.Common.Name + Year.of.Observation ~ Findings.of.Human.Interaction, value.var = 'cnt')
HI_yes_no_sp[is.na(HI_yes_no_sp)] <- 0
HI_yes_no_sp <- HI_yes_no_sp %>%
  transform(Yes = Y, No = (N+CBD)) %>%
  dplyr::select(Pinniped.Common.Name, Year.of.Observation, Yes, No) 

HI_yes_no_sp_period <- HI_yes_no_sp %>%
  transform(period = ifelse(Year.of.Observation < 2000, '1991-1999',
                            ifelse(Year.of.Observation < 2010 & Year.of.Observation > 1999, '2000-2009', '2009-2016'))) %>%
  group_by(period, Pinniped.Common.Name) %>%
  summarize(Yes = sum(Yes), No = sum(No)) %>%
  transform(HI_perc = Yes/(Yes+No))

#per year
# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'Harbor seal'))) 

#per period
summary(glm(cbind(Yes, No) ~ period, family = binomial(link = logit), data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'Harbor seal'))) #higher late: p < 0.001, z = 6.6, y = 0.6x

# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'California sea lion')))

summary(glm(cbind(Yes, No) ~ period, family = binomial, data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'California sea lion'))) #lower in middle, p < 0.001, z = -4.4, y = -0.7x

# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) #significant

summary(glm(cbind(Yes, No) ~ period, family = binomial, data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) 

# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'Northern elephant seal')))

summary(glm(cbind(Yes, No) ~ period, family = binomial, data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'Northern elephant seal'))) #late is almost high, p<0.1, z = 1.8, y = 0.8x

# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) #not sig

summary(glm(cbind(Yes, No) ~ period, family = binomial, data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'))) #not sig

# summary(glm(cbind(Yes, No) ~ Year.of.Observation, family = binomial, data = HI_yes_no_sp %>% filter(Pinniped.Common.Name == 'Northern fur seal'))) #not sig

summary(glm(cbind(Yes, No) ~ period, family = binomial, data = HI_yes_no_sp_period %>% filter(Pinniped.Common.Name == 'Northern fur seal'))) #not significant

# (c) Separate HI types, combined species
numberperyear <- pinnipeds_data %>% 
  group_by(Year.of.Observation) %>%
  summarize(yr.cnt = n_distinct(National.Database.Number))

HI_type_yes_no <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Interaction.Type, Year.of.Observation) %>%
  summarize(Yes = n_distinct(National.Database.Number)) %>%
  dcast(Year.of.Observation ~ Interaction.Type, value.var = 'Yes') 
HI_type_yes_no[is.na(HI_type_yes_no)] <- 0
HI_type_yes_no <- HI_type_yes_no %>%
  merge(numberperyear, by = c('Year.of.Observation')) %>%
  transform(NoBoat = yr.cnt - Boat, 
            NoFish = yr.cnt - Fisheries,
            NoGun = yr.cnt - Gunshot,
            NoOther = yr.cnt - Other) %>%
  dplyr::select(-c(yr.cnt)) 

HI_type_yes_no_per <- HI_type_yes_no %>% 
  transform(period = ifelse(Year.of.Observation < 2000, '1991-1999', 
              ifelse(Year.of.Observation < 2010 & Year.of.Observation > 1999, '2000-2009', '2009-2016'))) %>%
  group_by(period) %>%
  summarize(Boat = sum(Boat), NoBoat = sum(NoBoat),
            Fisheries = sum(Fisheries), NoFish = sum(NoFish),
            Gunshot = sum(Gunshot), NoGun = sum(NoGun),
            Other = sum(Other), NoOther = sum(NoOther))

#by period

#significantly higher late: p < 0.05, z = 2.4, y = 0.34x
summary(glm(cbind(Fisheries, NoFish) ~ period, family = binomial, data = HI_type_yes_no_per))
#significantly higher in both periods: p < 0.001, z = 3.4, y = 1.7x
summary(glm(cbind(Boat, NoBoat) ~ period, family = binomial, data = HI_type_yes_no_per))
#sig high late: p < 0.01, z = 3.1, y = 0.3x
summary(glm(cbind(Gunshot, NoGun) ~ period, family = binomial, data = HI_type_yes_no_per))
#sig higher in both: p < 0.001, z = 8.7, y = 1.2x
summary(glm(cbind(Other, NoOther) ~ period, family = binomial, data = HI_type_yes_no_per))

##yearly
# <-- Proportion of each interaction type changing over time, boat and other increasing at a faster rate than fisheries and gunshot wounds?
summary(glm(cbind(Boat, NoBoat) ~ Year.of.Observation, family = binomial, data = HI_type_yes_no)) # y = 0.08687x
summary(glm(cbind(Fisheries, NoFish) ~ Year.of.Observation, family = binomial, data = HI_type_yes_no)) # y = 0.043703x
summary(glm(cbind(Gunshot, NoGun) ~ Year.of.Observation, family = binomial, data = HI_type_yes_no)) # y = exp(0.039296*x) = 1.04^x
summary(glm(cbind(Other, NoOther) ~ Year.of.Observation, family = binomial, data = HI_type_yes_no)) # y = exp(0.08310*x) = 1.086^x

# (d) Separate HI types, separate species
numberperyearperspecies <- pinnipeds_data %>% 
  group_by(Year.of.Observation, Pinniped.Common.Name) %>%
  summarize(yr.cnt = n_distinct(National.Database.Number))

HIcount_type_species <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y' & 
           Pinniped.Common.Name != "Unidentified" &
           Interaction.Type != 'NA') %>%
  group_by(Interaction.Type, Year.of.Observation, Pinniped.Common.Name) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  dcast(Year.of.Observation + Pinniped.Common.Name ~ Interaction.Type, value.var = 'cnt')
HIcount_type_species[is.na(HIcount_type_species)] <- 0
HIcount_type_species <- HIcount_type_species %>%
  merge(numberperyearperspecies, by = c('Year.of.Observation', 'Pinniped.Common.Name')) %>%
  transform(NoBoat = yr.cnt - Boat, 
            NoFish = yr.cnt - Fisheries,
            NoGun = yr.cnt - Gunshot,
            NoOther = yr.cnt - Other) %>%
  dplyr::select(-c(yr.cnt))  

#(e) Which types of cases might be increasing or decreasing for which species?
#Not sig or insuff data for CSL, SSL, NFS, NES, GFS
summary(glm(cbind(Boat, NoBoat) ~ Year.of.Observation, family = binomial, data = HIcount_type_species %>% filter(Pinniped.Common.Name == 'Steller sea lion')))

#not sig for SSL, NFS, NES, GFS
summary(glm(cbind(Fisheries, NoFish) ~ Year.of.Observation, family = binomial, data = HIcount_type_species %>% filter(Pinniped.Common.Name == 'California sea lion')))

#not sig for harbors, SSL, NFS, NES, GFS
summary(glm(cbind(Gunshot, NoGun) ~ Year.of.Observation, family = binomial, data = HIcount_type_species %>% filter(Pinniped.Common.Name == 'California sea lion')))

#sig for CSL, HS, 
summary(glm(cbind(Other, NoOther) ~ Year.of.Observation, family = binomial, data = HIcount_type_species %>% filter(Pinniped.Common.Name == 'Harbor seal')))

```

## Seasonality  

- Is there a seasonal peak in strandings and HI cases for certain species? 
```{r}

#(a) Seasonal peak for overall counts?
 
monthly_all <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  group_by(Month.of.Observation, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <-- Yes, May through October is different from the rest of the year
kruskal.test(monthly_all)
kruskalmc(monthly_all$cnt ~ monthly_all$Month.of.Observation)
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_all, dist = "Chisquare")
  
#Alternative parametric pairwise method using glht - didn't explore too much
monthly_all_glht <- glm.nb(cnt ~ Month.of.Observation, data = monthly_all)
summary(glht(monthly_all_glht, linfct = mcp(Month.of.Observation = "Tukey")))


# (b) Is there a seasonal peak for certain species?

monthly_species <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Pinniped.Common.Name != 'Unidentified') %>%
  group_by(Month.of.Observation, Pinniped.Common.Name, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# Pairwise; repeat for each species
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'Harbor seal'), dist = "Chisquare") #June - September
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'California sea lion'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'Steller sea lion'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'Northern elephant seal'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'Northern fur seal'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'), dist = "Chisquare") #not significant; no strandings Jan-Mar; Dec
  
#Alternative parametetric pairwise method using glht
monthly_species_glht <- glm.nb(cnt ~ Month.of.Observation, data = monthly_species %>% filter(Pinniped.Common.Name == 'California sea lion'))
summary(glht(monthly_species_glht, linfct = mcp(Month.of.Observation = "Tukey")))

# #glm with and without interaction, anova the two
# #help, not sure how to interpret the anova output
# monthly_all_model_sp <- glm.nb(cnt ~ Pinniped.Common.Name + Month.of.Observation, data = monthly_species)
# monthly_all_model_sp_int <- glm.nb(cnt ~ Pinniped.Common.Name * Month.of.Observation, data = monthly_species)
# anova(monthly_all_model_sp, monthly_all_model_sp_int, test = 'Chisq')

```

```{r}

# Seasonality in HI cases?

#(a) All species combined
monthly_species_HI <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Pinniped.Common.Name != 'Unidentified') %>%
  filter(Findings.of.Human.Interaction == 'Y') %>%
  group_by(Month.of.Observation, Pinniped.Common.Name, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

#(b) separate species
# Pairwise; repeat for each species
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'Harbor seal'), dist = "Chisquare") #July
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'California sea lion'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'Steller sea lion'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'Northern elephant seal'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'Northern fur seal'), dist = "Chisquare") # not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_species_HI %>% filter(Pinniped.Common.Name == 'Guadalupe fur seal'), dist = "Chisquare") #not sig

# #Combined species; same as above, help, not sure how to interpret anova output
# monthly_HI_model_sp <- glm(cnt ~ Pinniped.Common.Name + Month.of.Observation, data = monthly_species_HI)
# monthly_HI_model_sp_int <- glm(cnt ~ Pinniped.Common.Name * Month.of.Observation, data = monthly_species_HI)
# anova(monthly_HI_model_sp, monthly_HI_model_sp_int, test = 'Chisq')

#(c) separate HI types?
monthly_HI_type <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  filter(Water.Body == 'Inland_WA') %>%
  filter(Pinniped.Common.Name == 'Harbor seal') %>%
  group_by(Month.of.Observation, Interaction.Type, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_HI_type %>% filter(Interaction.Type == 'Other'), dist = "Chisquare") 
```

- Seasonality across age and sex classes?
```{r}

#(a) Do certain age classes exhibit seasonality?

#age class across all strands by month? Help uncertain if I should have year here
monthly_age <- pinnipeds_data %>%
  filter(Year.of.Observation > 2001) %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Age.Class != 'Unid') %>%
  group_by(Month.of.Observation, Age.Class, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

#General model - at least one age class shows month as a predictor
age_month_model <- glm.nb(cnt ~ Age.Class + Month.of.Observation, data = monthly_age)
age_month_model_int <- glm.nb(cnt ~ Age.Class * Month.of.Observation, data = monthly_age)
anova(age_month_model, age_month_model_int, test = 'Chisq')

#Repeat for each age class
kruskal.test(monthly_age %>% filter(Age.Class == 'Pup'))
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_age %>% filter(Age.Class == 'Pup'), dist = "Chisquare") #June-Sept
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_age %>% filter(Age.Class == 'Yearling'), dist = "Chisquare") #not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_age %>% filter(Age.Class == 'Subadult'), dist = "Chisquare") #not sig
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, data = monthly_age %>% filter(Age.Class == 'Adult'), dist = "Chisquare") #not sig

#(b) Sex and seasonality?
Sex_all_mo <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Sex == 'Male' | Sex == 'Female') %>%
  group_by(Sex, Month.of.Observation, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

sex_month_model_m <- glm.nb(cnt ~ Month.of.Observation, data = Sex_all_mo %>% filter(Sex == 'Male'))
sex_month_model_f <- glm.nb(cnt ~ Month.of.Observation, data = Sex_all_mo %>% filter(Sex == 'Female'))
anova(sex_month_model_m, sex_month_model_f, test = 'Chisq')

#Repeat for each sex
posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, Sex_all_mo %>% filter(Sex == 'Male'), dist = "Chisquare") #July only different from Jan-April

posthoc.kruskal.nemenyi.test(cnt ~ Month.of.Observation, Sex_all_mo %>% filter(Sex == 'Female'), dist = "Chisquare") #July different from Jan-May and different from Nov-Dec
```

- Seasonality in HI prevalence of certain types?
```{r}

#Certain types of HI cases show seasonality?

numberpermonth <- pinnipeds_data %>% 
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  group_by(Month.of.Observation) %>%
  summarize(mo.cnt = n_distinct(National.Database.Number))

HI_type_yes_no_month <- pinnipeds_data %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Interaction.Type, Month.of.Observation) %>%
  summarize(Yes = n_distinct(National.Database.Number)) %>%
  dcast(Month.of.Observation ~ Interaction.Type, value.var = 'Yes') 
HI_type_yes_no_month[is.na(HI_type_yes_no_month)] <- 0
HI_type_yes_no_month <- HI_type_yes_no_month %>%
  merge(numberpermonth, by = c('Month.of.Observation')) %>%
  transform(NoBoat = mo.cnt - Boat, 
            NoFish = mo.cnt - Fisheries,
            NoGun = mo.cnt - Gunshot,
            NoOther = mo.cnt - Other) %>%
  dplyr::select(-c(mo.cnt)) %>%
  transform(Boat.prop = (Boat/NoBoat),
            Fisheries.prop = (Fisheries/NoFish),
            Gunshot.prop = (Gunshot/NoGun),
            Other.prop = (Other/NoOther))

#Proportion of types - for more descriptive analysis?
numberperyearpermonth <- pinnipeds_data %>% 
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  group_by(Month.of.Observation, Year.of.Observation) %>%
  summarize(mo.cnt = n_distinct(National.Database.Number))

HIprev_type_mo <- pinnipeds_data %>%
  filter(Water.Body == 'WA_Coast') %>%
  filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Interaction.Type, Month.of.Observation, Interaction.Type, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  dcast(Month.of.Observation + Year.of.Observation ~ Interaction.Type, value.var = 'cnt')
HIprev_type_mo[is.na(HIprev_type_mo)] <- 0
HIprev_type_mo <- HIprev_type_mo %>%
  merge(numberperyearpermonth, by = c('Month.of.Observation', 'Year.of.Observation')) %>%
  transform(Boat = (Boat/mo.cnt),
            Fisheries = (Fisheries/mo.cnt),
            Gunshot = (Gunshot/mo.cnt),
            Other = (Other/mo.cnt)) %>%
  dplyr::select(Month.of.Observation, Year.of.Observation, Boat, Gunshot, Fisheries, Other)

HIprev_type_mo_sub <- pinnipeds_data %>%
  #filter(Observation.Status == 'ALIVE' | Observation.Status == 'FRESH DEAD') %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Interaction.Type, Month.of.Observation, Water.Body, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  filter(Water.Body == 'OR_Coast' & Interaction.Type == 'Gunshot') %>% group_by(Month.of.Observation) %>% dplyr::summarize(cnt = sum(cnt))
#   dcast(Month.of.Observation + Year.of.Observation + Water.Body ~ Interaction.Type, value.var = 'cnt')
HIprev_type_mo_sub[is.na(HIprev_type_mo_sub)] <- 0

ggplot(HIprev_type_mo_sub, aes(y = cnt, x = factor(Month.of.Observation))) +
         geom_point()

#Very prevalent gunshot wounds in March - fisheries season opening?
marchgun <- HIprev_type_mo %>% filter(Month.of.Observation == 'MAR') %>%
  summarize(mean(Gunshot))

posthoc.kruskal.nemenyi.test(Other ~ Month.of.Observation, HIprev_type_mo, dist = "Chisquare") 

```


## Spatial Patterns  

- Is there a statistical difference in strandings and HI across regions/counties over time? (Washington)
```{r}

# (a) Strandings in 3 sub-regions changing over time?
region_yr <- pinnipeds_data %>%
  filter(!is.na(Water.Body) & Water.Body != 'NA') %>% 
  group_by(Water.Body, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

summary(glm.nb(cnt ~ Year.of.Observation, data = region_yr %>% filter(Water.Body == 'Inland_WA')))
#y = exp(0.03984) = 1.040^x, z = 4.95, p < 0.001

summary(glm.nb(cnt ~ Year.of.Observation, data = region_yr %>% filter(Water.Body == 'WA_Coast')))
#y = exp(0.0714) = 1.074^x, z = 3.62, p < 0.001

summary(glm.nb(cnt ~ Year.of.Observation, data = region_yr %>% filter(Water.Body == 'OR_Coast')))
#y = exp(0.05044) = 1.052^x, z = 3.78, p < 0.001

#County level:

#(a) All counties all strandings
bycounty <- pinnipeds_data %>% 
  filter(County != 'NA' & County != 'Unknown') %>%
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

kruskal.test(bycounty)

#All strandings, WA
allWA_bycounty <- pinnipeds_data %>% 
  filter(County != 'NA' & County != 'Unknown' & State == 'WA') %>%
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <--- Overall, yes, county is a significant predictor. 
kruskal.test(allWA_bycounty)

# <--- Pairwise shows San Juan dif from Mason, Pacific, Clark, Cowlitz, Thurston
#kruskalmc(allWA_bycounty$cnt ~ allWA_bycounty$County) 
posthoc.kruskal.nemenyi.test(cnt ~ County, allWA_bycounty, dist = "Chisquare")

# (b) HI strandings, WA
WAHI_bycounty <- pinnipeds_data %>% 
  filter(County != 'NA' & State == 'WA' & 
           Findings.of.Human.Interaction == 'Y' & 
           Interaction.Type != 'NA') %>% 
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <--- County is a significant predictor overall. 
kruskal.test(WAHI_bycounty)

# <--- Pairwise shows higher in Pacific
kruskalmc(WAHI_bycounty$cnt ~ WAHI_bycounty$County)
posthoc.kruskal.nemenyi.test(cnt ~ County, WAHI_bycounty, dist = "Chisquare")


```

- Is there a statistical difference in strandings and HI across counties? (both states combined)
```{r}

#(a) All strandings, both states
all_bycounty <- pinnipeds_data %>% 
  filter(County != 'NA' & County != 'Unknown') %>% 
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

countylist <- all_bycounty %>% group_by(County) %>%
  summarize(cnt = sum(cnt))

# Would be best to have the control be the mean - trying "contr.sum"
county_contrasts <- summary(glm(cnt ~ County, contrasts = list(County = 'contr.sum'), data = all_bycounty))

# <--- K-W show strandings higher in Lincoln
# kruskal.test(allOR_bycounty)
# posthoc.kruskal.nemenyi.test(cnt ~ County, allOR_bycounty, dist = "Chisquare")

# (b) HI strandings
HI_bycounty <- pinnipeds_data %>% 
  filter(County != 'NA' & 
           Findings.of.Human.Interaction == 'Y' & 
           Interaction.Type != 'NA') %>% 
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

county_contrasts_HI <- summary(glm(cnt ~ County, contrasts = list(County = 'contr.sum'), data = HI_bycounty))

# <--- K-W shows Clatsop, Lincoln
# kruskal.test(ORHI_bycounty)
# posthoc.kruskal.nemenyi.test(cnt ~ County, ORHI_bycounty, dist = "Tukey")


```

Is the prevalence of HI cases higher than expected given the distribution of overall strandings?
```{r}

#Is the prevalence of HI cases higher than expected given the distribution of overall strandings?
#Help - never quite got this?

# Yes, No for HI by county
HI_yes_no_county <- pinnipeds_data %>%
  filter(County != 'Unknown') %>%
  group_by(County, Findings.of.Human.Interaction) %>%
  summarize(cnt = n_distinct(National.Database.Number)) %>%
  dcast(County ~ Findings.of.Human.Interaction, value.var = 'cnt')
HI_yes_no_county[is.na(HI_yes_no_county)] <- 0
HI_yes_no_county <- HI_yes_no_county %>%
  transform(Yes = Y, No = (N+CBD)) %>%
  dplyr::select(County, Yes, No) 


```


- Are specific types of HI cases increasing or decreasing in certain counties?
```{r}

# Are gunshot, entanglements, or boat collisions increasing in certain counties? 

HI_type_yr <- pinnipeds_data %>% 
  filter(County != 'NA' & 
           Findings.of.Human.Interaction == 'Y' & 
           Interaction.Type != 'NA') %>% 
  transform(County = factor(County)) %>%
  group_by(County, Year.of.Observation, Interaction.Type) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <--- Shows increasing in Grays, Pacific, Pierce, CLallam, Whatcom?
HI_type_yr_model <- glm.nb(cnt ~ Year.of.Observation, HI_type_yr %>% filter(County == 'Pacific'))
summary(HI_type_yr_model)
# Pierce: y = exp(0.06673) = 6.9%; z = 3.03, p < 0.01
# Pacific: y = 7.3%; z = 2.7, p < 0.01
# Clatsop: y = 1.095^x = 9.5%, 3.9, p < 0.001
# Whatcom: y = 0.06371^x = 1.066, 3.0, p < 0.001
# King: y = 0.05124^x = 1.053, 2.74, p < 0.01
# Jefferson: y = 0.0736^x = 1.076, 2.2, p < 0.05
# Lincoln; Grays Harbor; San Juan; Island; Snohomish; Skagit; Thurston; Tilla; Coos: not sig

# <--- Shows only "other" increasing in Pierce?
WAHI_type_yr_model <- glm(cnt ~ County*Interaction.Type, family = poisson(link = log), WAHI_type_yr)
summary(WAHI_type_yr_model)


```


## Categorical/Basic Species, Age and Sex Classes  

- Is there a difference in species of strandings and HI cases? 

```{r}

#(a) all strandings
species <- pinnipeds_data %>%
  group_by(Pinniped.Common.Name) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <--- Yes, species are different
kruskal.test(species)

#(b) HI cases
species_HI <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y') %>%
  group_by(Pinniped.Common.Name) %>%
  summarize(cnt = n_distinct(National.Database.Number))

# <--- Yes, species are different
kruskal.test(species_HI)
kruskalmc(species_HI$cnt ~ species_HI$Pinniped.Common.Name)

```

- Is there a difference in sex of strandings and HI cases?
```{r}

# Is there a difference in sex of strandings/HI cases?  
 
#(a) All strandings
Sex_all_yr <- pinnipeds_data %>%
  filter(Pinniped.Common.Name == 'Steller sea lion') %>%
  filter(Sex == 'Male' | Sex == 'Female') %>%
  group_by(Sex, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# <--- Yes, males different (higher) from females
kruskal.test(Sex_all_yr)
kruskalmc(Sex_all_yr$cnt ~ Sex_all_yr$Sex)

#(b) HI strandings
Sex_HI_yr <- pinnipeds_data %>%
  filter(Sex == 'Male' | Sex == 'Female') %>%
  filter(Findings.of.Human.Interaction == 'Y' & Interaction.Type != 'NA') %>%
  group_by(Sex, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# <-- Yes, males different (higher) from females
kruskal.test(Sex_HI_yr)
kruskalmc(Sex_HI_yr$cnt ~ Sex_HI_yr$Sex)
# sex_HI_model <- glm.nb(cnt ~ Sex, Sex_HI_yr)
# summary(sex_HI_model)

```


- Is there a difference in age class of strandings and HI cases?
```{r}
# Is there a difference in age class strandings/HI cases? 

#(a) all strandings
age_all <- pinnipeds_data %>%
  filter(Year.of.Observation > 2001) %>%
  group_by(Age.Class, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# <--- Yes, pups and adults (not different from each other) are different from subadults and yearlings (not different from each other). p < 0.05, difs amount to ~38 animals/yr
kruskal.test(age_all)
kruskalmc(age_all$cnt ~ age_all$Age.Class)

#(b) HI strandings
age_HI <- pinnipeds_data %>%
  filter(Findings.of.Human.Interaction == 'Y') %>%
  group_by(Age.Class, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number)) 

# <--- Yes, yearlings significantly different (lower). p < 0.05, difs amount to ~38 animals/yr
kruskal.test(age_HI)
kruskalmc(age_HI$cnt ~ age_HI$Age.Class)

#contrast for difference from mean rather than reference point
age_contrasts_HI <- summary(glm(cnt ~ Age.Class, contrasts = list(Age.Class = 'contr.sum'), data = age_HI))

#paper question - proportion of pups that are harbors?

hs_pups <- pinnipeds_data %>%
  filter(Pinniped.Common.Name == 'Harbor seal' & Age.Class == 'Pup') %>%
  summarize(cnt = n_distinct(National.Database.Number))

all_pups <- pinnipeds_data %>%
  filter(Age.Class == 'Pup') %>%
  summarize(cnt = n_distinct(National.Database.Number))

hs_pups/all_pups


```

- Is there a difference between ages and sexes?
```{r }

# Help: would like non-parametric test for multiple categorical variables

#(a) Differences between ages and sexes?  
age_sex_all <- pinnipeds_data %>%
  group_by(Sex, Age.Class, Year.of.Observation) %>%
  summarize(cnt = n_distinct(National.Database.Number))

#Need non-parametric version of this, for multiple comparisons :(  
TukeyHSD(aov(cnt ~ Sex*Age.Class, data = age_sex_all))

# kruskalmc(age_sex_all$cnt ~ age_sex_all$Age.Class*age_sex_all$Sex)
# posthoc.kruskal.nemenyi.test(cnt ~ Age.Class + Sex, age_sex_all, dist = "Chisquare")

```

