---
title: 'Pinniped Stranding and Oceanographic Index Modeling Analysis and Research Questions'
author: Amanda Warlick, March 2017
output:
  word_document:
    reference_docx: mytemplate.docx

---

```{r, include = F}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
```
 

```{r, include = F, echo = F}
library(ggplot2)
library(tidyr)
library(ggmap)
library(data.table)
library(cowplot)
library(knitr)
library(reshape2)
library(stringr)
library(magrittr)
library(scales)
library(stats) 
library(PMCMR)
library(pgirmess) #kruskalmc()
library(MASS) #glm.nb()
library(multcomp) #glht()
library(dplyr)
# detach("package:multcomp")
# detach("package:TH.data")
# detach("package:MASS")

#setwd("~/Documents/R/Strandings")

# Data Load 
###############################################################################
pinnipeds_ocean <- read.csv("pinnipeds_ocean.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  # transform(Decade = ifelse(Year.of.Observation < 2000, '1990s', 
  #                           ifelse(Year.of.Observation < 2010, '2000s', '2010s'))) %>%
  transform(Period = ifelse(Year.of.Observation < 2005, 'Pre_Prescott', 'Post_Prescott'))

# Set Up Factors
# pinnipeds_data$Age.Class <- factor(pinnipeds_data$Age.Class, 
#                                    levels = c("Pup", "Yearling", "Subadult", "Adult", "Unid"))

pinnipeds_ocean_small <- pinnipeds_ocean %>%
  
```

##Methodological Questions

- Appropriate to apply breakpoint and then two different means for more even anomalies over time?
- What are the things I should be cognizant of that are particular to my data? 
- How to deal with colinearity of oceanographic conditions, and seasonality of all strandings?
- Is applying a timeseries monthly average (all Januaries) to each month the appropriate way to derive monthly anomalies?
- How will the models deal with certain oceanographic variables not being available for the whole time series?

##Research Questions

1. Are strandings (and or HI cases) for combined or individual species correlated with certain oceanographic indices?

1a. All cases
```{r}

# (a) Strandings for combined species 

# (b) Strandings for separate species

```

1b. HI cases
```{r}
# (a) HI cases for combined species 

# (b) HI cases for separate species

```

2. Are stranding anomalies correlated with certain oceanographic indices?

2a. All cases
```{r}

# (a) Stranding anoms for combined species 

# (b) Stranding anoms for separate species

```

2b. HI cases
```{r}

# (a) HI case anoms for combined species 

# (b) HI case anoms for separate species

```

3. Is there a detectable and appropriate lag time where oceanographic indices correlate more strongly with strandings or stranding anomalies?

3a. 
```{r}

#All strandings combined species

#All stranding anomalies combined species

#All HI cases combined species

#all HI anomalies combined species

#Separate species?
```

4. Are there different models that are a better fit for different species (i.e., are species impacted differently by the same ocean conditions?)
```{r}

```

5. Are stranding hotspots different during EN vs LN conditions? Or some other metric?
```{r}

# (a) All strandings

# (b) HI cases combined

# (c) HI case types
```


