---
title: 'Pinniped Stranding and Oceanographic Index Modeling Analysis and Research Questions'
author: Amanda Warlick, March 2017
output:
  word_document

---

```{r, include = F}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
```
 

```{r packages and data, include = F, echo = F}
library(ggplot2)
library(tidyr)
#library(ggmap)
library(data.table)
#library(cowplot)
library(knitr)
library(reshape2)
library(stringr)
library(magrittr)
library(scales)
library(stats) 
library(MASS) #glm.nb(), stepAIC()
library(dplyr)
library(leaps) # regsubsets() 
library(corrplot)
library(Hmisc) # rcorr()

#setwd("~/Documents/R/Strandings/ENSO")

# Data Load 
###############################################################################
pinnipeds_ocean <- read.csv("pinnipeds_ocean.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2005, 'Pre_Prescott', 'Post_Prescott'))

cetaceans_ocean <- read.csv("cetaceans_ocean.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2005, 'Pre_Prescott', 'Post_Prescott'))

#Load dataframes from PinnipedsandENSO.Rmd

#Ocean only variables for correlation exploration
ocean_corr <- read.csv("ocean_full.csv", header = T, na.strings = "NA", stringsAsFactors = F) %>%
  dplyr::select(-matches("Date")) %>%
  dplyr::select(-matches("ENSO_cat")) %>%
  dplyr::select(-c(Month.of.Observation, Year.of.Observation))

#Monthly anomalies separate species
monthly_anom <- read.csv("monthly_anom.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2004, 'Pre_Prescott', 'Post_Prescott'))
#Monthly anomalies combined species
monthly_anom_comb <- read.csv("monthly_anom_comb.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2004, 'Pre_Prescott', 'Post_Prescott'))
#Annual anomalies separate species
annual_anom <- read.csv("annual_anom.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2004, 'Pre_Prescott', 'Post_Prescott'))
#Annual anomalies combined species
annual_anom_comb <- read.csv("annual_anom_comb.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE) %>%
  transform(Period = ifelse(Year.of.Observation < 2004, 'Pre_Prescott', 'Post_Prescott'))

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
```

```{r ocean cor}

ocean_corr_small <- ocean_corr %>%
  dplyr::select(-matches("lag")) %>%
  dplyr::select(-matches("44")) %>%
  dplyr::select(-matches("45")) %>%
  rename(Upwelling = Upwelling_39, SST = SST_39, SST_anom = SST_39_anom, Wind = Wind_39) %>%
  dplyr::select(MEI, PDO, NPGO, SST, SST_anom, Upwelling, Wind, CBA_North, CBA_South)

ocean_corr_0vs1 <- ocean_corr %>%
  dplyr::select(-matches("44")) %>%
  dplyr::select(-matches("45")) %>%
  rename(Upwelling = Upwelling_39, SST = SST_39, SST_anom = SST_39_anom, Wind = Wind_39) %>%
  dplyr::select(-matches("lag2")) %>%
  dplyr::select(-matches("lag3")) 

ocean_corr_0vs2 <- ocean_corr %>%
  dplyr::select(-matches("lag1")) %>%
  dplyr::select(-matches("lag3")) 

ocean_corr_0vs3 <- ocean_corr %>%
  dplyr::select(-matches("lag1")) %>%
  dplyr::select(-matches("lag2")) 

ocean_corr_small_buoys <- ocean_corr %>%
  dplyr::select(-matches("lag")) 

# cor <- cor(ocean_corr_small, use = "pairwise.complete.obs")
# cor <- cor(ocean_corr_med, use = "pairwise.complete.obs")

cor_test_small <- rcorr(as.matrix(ocean_corr_small))
flattenCorrMatrix(cor_test_small$r, cor_test_small$P)

cor_test_small_buoys <- rcorr(as.matrix(ocean_corr_small_buoys))

cor_test_0vs1 <- rcorr(as.matrix(ocean_corr_0vs1))
cor_test_0vs2 <- rcorr(as.matrix(ocean_corr_0vs2))
cor_test_0vs3 <- rcorr(as.matrix(ocean_corr_0vs3))

#In general, 
## --- Upwelling 39 and 45 are correlated, likewise Wind 39 and 44, SST 39 and 44 less so
## --- SST and SST_anom of same buoy are correlated, less so for opposite buoy pairs
## --- Upwelling and Wind are inversely correlated
## --- PDO and MEI correlated
## --- CBA North and South inversely correlated, and correlated to MEI etc. and SST_anoms, but less so with SST, Upwell, and Wind
small <- corrplot(cor_test_small$r, type = "lower", order = "original", p.mat = cor_test_small$P, 
         insig = "blank", sig.level = 0.01, tl.col = "black", tl.cex = .9, number.cex = .2)
lag0vs1 <- corrplot(cor_test_0vs1$r, type = "lower", order = "original", p.mat = cor_test_0vs1$P, 
         insig = "blank", sig.level = 0.01, tl.col = "black", tl.cex = .7, number.cex = .7)
#SST and SST_anom_lag2s are correlated
lag1vs2 <- corrplot(cor_test_0vs2$r, type = "lower", order = "original", p.mat = cor_test_0vs2$P, 
         insig = "blank", sig.level = 0.01, tl.col = "black", tl.cex = .7, number.cex = .7)
lag2vs3 <- corrplot(cor_test_0vs3$r, type = "lower", order = "original", p.mat = cor_test_0vs3$P, 
         insig = "blank", sig.level = 0.01, tl.col = "black", tl.cex = .7, number.cex = .7)

corrplot(cor_test_small_buoys$r, type = "lower", order = "original", p.mat = cor_test_small_buoys$P, 
         insig = "blank", sig.level = 0.001, tl.col = "black", tl.cex = .9, number.cex = .2)

```

```{r chow test}

r.reg <- lm(cnt_all ~ Year.of.Observation, data = annual_anom_comb)
ur.reg1 <- lm(cnt_all ~ Year.of.Observation, data = annual_anom_comb %>% filter(Period == 'Pre_Prescott'))
ur.reg2 <- lm(cnt_all ~ Year.of.Observation, data = annual_anom_comb %>% filter(Period != 'Pre_Prescott'))
                                            
## review the regression results
summary(r.reg)
summary(ur.reg1)
summary(ur.reg2)
 
## Calculate sum of squared residuals for each regression
SSR <- NULL
SSR$r <- r.reg$residuals^2
SSR$ur1 <- ur.reg1$residuals^2
SSR$ur2 <- ur.reg2$residuals^2
 
## K is the number of regressors in our model
K <- r.reg$rank
 
## Computing the Chow test statistic (F-test)
numerator <- (sum(SSR$r) - (sum(SSR$ur1) + sum(SSR$ur2)))/K
denominator <- (sum(SSR$ur1) + sum(SSR$ur2)) / (nrow(annual_anom_comb) - 2*K)
chow = numerator/denominator
chow
 
## Calculate P-value
1-pf(chow, K, (nrow(annual_anom_comb) - 2*K))

## Plot the results
ggplot(annual_anom_comb, aes(Year.of.Observation, cnt_all)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F, col = 'grey50', linetype = 'dashed', size = 0.7) +
  geom_smooth(aes(group = Period), method = 'lm', se = F, linetype = 'dashed', col = 'red', size = 0.7) + 
  xlab("") + ylab("Total Strandings")

ggplot(annual_anom_comb, aes(Year.of.Observation, cnt_HI)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F, col = 'grey50', linetype = 'dashed', size = 0.7) +
  geom_smooth(aes(group = Period), method = 'lm', se = F, linetype = 'dashed', col = 'red', size = 0.7) + 
  xlab("") + ylab("HI Cases")

#Strucchange option - comes up with same pvalue as manual calculation, 2003 breakpoint
sctest(cnt_all ~ Year.of.Observation, type = "Chow", point = 15, data = annual_anom_comb)
sctest(strand_anom ~ Year.of.Observation, type = "Chow", point = 15, data = annual_anom_comb)

sctest(cnt_HI ~ Year.of.Observation, type = "Chow", point = 15, data = annual_anom_comb)
sctest(strand_anom_HI ~ Year.of.Observation, type = "Chow", point = 15, data = annual_anom_comb)
```

##Research Questions

1. Are strandings (and or human interaction (HI) cases) for combined or individual species correlated with certain oceanographic indices?

1a. All cases, plus lag variables
```{r}

#All cases monthly

# (a) Strandings for combined species 

#Model selection using stepAIC() on negative binomial regression
#na.omit() removes all rows with any missing data, glm.nb(keeps rows but removes the single NA value)
#keep this - list of all vars
#keep this - all vars w/o 44 and lags: PDO + MEI + NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South +
          #CBA_North + Period
fit_step_all <- glm.nb(cnt_all ~ 
          MEI + PDO + NPGO + SST_39 + SST_39_anom + SST_44 + SST_44_anom + Wind_39 + Wind_44 + Upwelling_39 + Upwelling_45 + CBA_North + CBA_South +
          MEI_lag1 + PDO_lag1 + NPGO_lag1 + SST_39_lag1 + SST_44_lag1 + Wind_39_lag1 + Wind_44_lag1 + Upwelling_39_lag1 + Upwelling_45_lag1 + CBA_North_lag1 + CBA_South_lag1 + SST_39_anom_lag1 + SST_44_anom_lag1 +
          MEI_lag2 + PDO_lag2 + NPGO_lag2 + SST_39_lag2 + SST_44_lag2 + Wind_39_lag2 + Wind_44_lag2 + Upwelling_39_lag2 + Upwelling_45_lag2 + CBA_North_lag2 + CBA_South_lag2 + SST_39_anom_lag2 + SST_44_anom_lag2 +
          MEI_lag3 + PDO_lag3 + NPGO_lag3 + SST_39_lag3 + SST_44_lag3 + Wind_39_lag3 + Wind_44_lag3 + Upwelling_39_lag3 + Upwelling_45_lag3 + CBA_North_lag3 + CBA_South_lag3 + SST_39_anom_lag3 + SST_44_anom_lag3, 
      data = na.omit(monthly_anom_comb))
step <- stepAIC(fit_step_all, direction = "backward")
step$anova

#regsubsets() - chooses the nbest models for up to 8 predictors
mod_select <- regsubsets(cnt_all ~ 
      MEI + NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South +
          CBA_North + Period, data = monthly_anom_comb, nbest = 5)
summary(mod_select)

#best model without lag data was: SST_44, SST_39_anom, Upwelling_39, Wind_39
nb <- summary(glm.nb(cnt_all ~ MEI + NPGO + SST_39 + Wind_39 + CBA_South +
          Period, data = monthly_anom_comb))
qp <- summary(glm(cnt_all ~ SST_39_anom + SST_44 + Upwelling_39 + CBA_South + Wind_39, data = monthly_anom_comb, family = 'quasipoisson'))

#best model with lag data; #PDO and MEI seem interchangeable in the model
nb <- summary(glm.nb(cnt_all ~ SST_39_anom + SST_44 + CBA_South_lag1 + Wind_39_lag1 + 
                       Upwelling_39_lag1 + MEI_lag1 + Wind_39_lag2 + Upwelling_39_lag2 +  SST_39_lag3 + Period, 
                     data = monthly_anom_comb))

# qp <- summary(glm(cnt_all ~ SST_39_anom + SST_44 + CBA_South_lag1 + Wind_39_lag1 + Upwelling_39_lag1 + 
#                     MEI_lag1 + Wind_39_lag2 + Upwelling_39_lag2 +  SST_39_lag3 + Period, 
#                   data = monthly_anom_comb, family = 'quasipoisson'))

pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 0.33, which means model is an okay fit?
# pchisq(deviance(qp), df.residual(qp), lower.tail = F) # pvalue is small - not a good fit


# (b) Strandings for separate species

#California sea lion

#regsubsets, continue with SST44, MEI, SST39anom, Upwelling_39 OR Wind_44, CBA_South

mod_select <- regsubsets(cnt_all ~ PDO + MEI + NPGO + SST_39 + SST_39_anom + Wind_39 + CBA_South + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), nbest = 3)
summary(mod_select)

#wo lag
nb <- summary(glm.nb(cnt_all ~ NPGO + SST_39 + SST_39_anom + Period, 
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion')))
pchisq(deviance(nb), df.residual(nb), lower.tail = F) #.85


#w lag
mod_select <- regsubsets(cnt_all ~ PDO + NPGO + SST_39 + SST_39_anom + SST_44 + SST_44_anom + Wind_44 + CBA_South +
          SST_44_lag1 + SST_39_anom_lag1 +
          SST_39_lag2 + Wind_39_lag2 + Upwelling_39_lag2 + SST_39_anom_lag2 + 
          Wind_44_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), nbest = 3, really.big = T)
summary(mod_select)

#without lag variables, best fit was SST_44, SST_39_anom, and Wind_44 OR Upwelling_39 significant
nb <- summary(glm.nb(cnt_all ~ SST_39 + SST_39_anom + SST_39_anom_lag3 + SST_44 + SST_44_lag1 +
                 Wind_39_lag2 + Wind_44 + Wind_44_lag3 + CBA_South + Upwelling_39_lag2 + Period, 
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion')))

# qp <- summary(glm(cnt_all ~ SST_39 + SST_39_anom + SST_39_anom_lag3 + SST_44 + SST_44_lag1 +
#                  Wind_39_lag2 + Wind_44 + Wind_44_lag3 + CBA_South + Upwelling_39_lag2 + Period, 
#           data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), family = 'quasipoisson'))
pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 0.48, which means model is an okay fit?
# pchisq(deviance(qp), df.residual(qp), lower.tail = F) # not a good fit - overdispersion

#Harbor seal

#wo lag
mod_select <- regsubsets(cnt_all ~ PDO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 +
          CBA_North + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'), nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(cnt_all ~ SST_39 + SST_39_anom + Upwelling_39 +
          CBA_North + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'))) 
pchisq(deviance(nb), df.residual(nb), lower.tail = F)

#w lag
#regsubsets, continue with Wind_39, Upwelling_39, SST_39, SST_39_anom, PDO, SST_44
mod_select <- regsubsets(cnt_all ~ SST_39 + SST_39_anom + SST_44 + SST_44_anom + Wind_39 + CBA_South +
         SST_44_lag1 + SST_44_lag2 + Upwelling_45_lag2 + SST_44_lag3 + SST_44_anom_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'), nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(cnt_all ~ SST_44 + CBA_South_lag3 + Wind_39 + SST_44_lag3 + SST_44_anom_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'))) #neg binomial produces lower AIC
# qp <- summary(glm(cnt_all ~ SST_44 + CBA_South_lag3 + Wind_39 + SST_44_lag3 + SST_44_anom_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal')), family = 'quasipoisson')

pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 0.79, which means model is an okay fit?
# pchisq(deviance(qp), df.residual(qp), lower.tail = F) # pvalue < 0.05, indicates not a good fit

#Steller sea lion

#wo lags
mod_select <- regsubsets(cnt_all ~ PDO + MEI + NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South +
          Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion'), nbest = 3, really.big = T)
summary(mod_select)
#warning messages
nb <- summary(glm.nb(cnt_all ~ SST_39 + SST_39_anom + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion'))) #pchisq = 0.99
# qp <- summary(glm(cnt_all ~ SST_39 + SST_39_anom +
#           Period, family = 'quasipoisson', data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')))
  
#w lag
#regsubsets, continue with Wind_39, CBA_North, SST_39, SST_39_anom, MEI, SST_44
mod_select <- regsubsets(cnt_all ~ MEI + SST_44 + Wind_39 + Upwelling_39 + CBA_North +
          MEI_lag1 + SST_44_lag2 + Wind_39_lag2 + CBA_North_lag2 +
          MEI_lag3 + Wind_44_lag3 + Upwelling_45_lag3 + CBA_North_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion'), nbest = 3, really.big = T)
summary(mod_select)

#Warning message
nb <- summary(glm.nb(cnt_all ~ MEI + Wind_39 + CBA_North + MEI_lag1 + CBA_North_lag2 + MEI_lag3 + Upwelling_45_lag3 + CBA_North_lag3, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')))

qp <- summary(glm(cnt_all ~ MEI + Wind_39 + CBA_North + MEI_lag1 + CBA_North_lag2 + MEI_lag3 + Upwelling_45_lag3 + CBA_North_lag3 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')), family = 'quasipoisson') 

pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 0.99
pchisq(deviance(qp), df.residual(qp), lower.tail = F) # pvalue = .2

#<-- Help: Overall for these single-species models, it looks like the recommended best-fit models are different from each other, but by how much or whether it is meaningful is unclear to me
#<-- Help: For all of the above, what to do about colinearity/covariance of oceanographic variables?

```

1b. HI cases
```{r}

#no lags or 44: #keep this - all vars w/o 44 and lags: PDO + MEI + NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South + CBA_North + Period

# (a) HI cases for combined species 

#wo lag
mod_select <- regsubsets(cnt_HI ~ NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South + CBA_North + Period, data = monthly_anom_comb, nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(cnt_HI ~ SST_39 + SST_39_anom + Wind_39 + CBA_South + Period, 
          data = monthly_anom_comb))
qp <- summary(glm(cnt_HI ~ NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South + CBA_North + Period, 
          data = monthly_anom_comb, family = 'quasipoisson'))
pchisq(deviance(nb), df.residual(nb), lower.tail = F)

#w lag
mod_select <- regsubsets(cnt_HI ~ MEI + SST_39 + Wind_44 + CBA_South + SST_39_lag1 + Wind_39_lag3 + Period, data = monthly_anom_comb, nbest = 3, really.big = T)
summary(mod_select)

qp <- summary(glm(cnt_HI ~ MEI + SST_39 + CBA_South + SST_39_lag3 + Period, 
          data = monthly_anom_comb, family = 'quasipoisson'))
pchisq(deviance(qp), df.residual(qp), lower.tail = F) #high pvalue

# (b) HI cases for separate species

#California sea lion

#wo lags
mod_select <- regsubsets(cnt_HI ~ MEI + SST_39 + Upwelling_39 + Wind_39 + CBA_South + CBA_North + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(cnt_HI ~ PDO + SST_39 + Wind_44 + CBA_South + Upwelling_45_lag2,
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion')))

qp <- summary(glm(cnt_HI ~ PDO + SST_39 + Wind_44 + CBA_South + Upwelling_45_lag2, 
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), family = 'quasipoisson'))
pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 
pchisq(deviance(qp), df.residual(qp), lower.tail = F) 

#w lags
#regsubsets
mod_select <- regsubsets(cnt_HI ~ PDO + SST_39 + SST_39_anom + Wind_44 + CBA_North + CBA_South +
          MEI_lag1 + PDO_lag1 + NPGO_lag1 + Wind_39_lag1 + Upwelling_45_lag2 + CBA_South_lag2 + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(cnt_HI ~ PDO + SST_39 + Wind_44 + CBA_South +
          Upwelling_45_lag2,
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion')))

qp <- summary(glm(cnt_HI ~ PDO + SST_39 + Wind_44 + CBA_South + Upwelling_45_lag2 + CBA_South_lag2, 
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'California sea lion'), family = 'quasipoisson'))
pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 0.99
pchisq(deviance(qp), df.residual(qp), lower.tail = F) # pvalue = 0.99, which means model is an okay fit?

#Harbor seal

#wo lags - nothing significant
# mod_select <- regsubsets(cnt_HI ~ PDO + MEI + NPGO + SST_39 + Upwelling_39 + Wind_39 + CBA_North + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'), nbest = 3, really.big = T)
# summary(mod_select)
# 
# nb <- summary(glm.nb(cnt_HI ~ NPGO + SST_39 + Upwelling_39 + CBA_North, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal')))
# qp <- summary(glm(cnt_HI ~ NPGO + SST_39 + Upwelling_39 + CBA_North, 
#           data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal')), family = 'quasipoisson')

#w lags
#regsubsets
mod_select <- regsubsets(cnt_HI ~
         PDO + SST_44 + Wind_39 + Upwelling_39 + CBA_North + CBA_South +
          PDO_lag1 + Upwelling_45_lag1 +  
          SST_44_lag2 + CBA_South_lag2 +
          SST_39_lag3 + Wind_44_lag3 + Upwelling_45_lag3 + CBA_South_lag3 + SST_39_anom_lag3 + SST_44_anom_lag3, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal'), nbest = 3, really.big = T)
summary(mod_select)

# nb <- summary(glm.nb(cnt_HI ~ PDO + SST_44 + Wind_39 + Upwelling_39 + CBA_North +
#           SST_44_lag2 + SST_39_anom_lag3, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal')))

qp <- summary(glm(cnt_HI ~ SST_44 + CBA_North +
          SST_44_lag2 + Upwelling_45_lag3 + SST_39_anom_lag3, 
          data = monthly_anom %>% filter(Pinniped.Common.Name == 'Harbor seal')), family = 'quasipoisson')

pchisq(deviance(nb), df.residual(nb), lower.tail = F) # pvalue = 1 - something is clearly wrong here
pchisq(deviance(qp), df.residual(qp), lower.tail = F) # pvalue < 0.99

#Steller sea lion

#wo lags
mod_select <- regsubsets(cnt_HI ~ MEI + NPGO + SST_39 + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South + Period, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion'), nbest = 3, really.big = T)
summary(mod_select)

qp <- summary(glm(cnt_HI ~ NPGO + SST_39_anom + Upwelling_39 + Wind_39 + CBA_South, 
                  family = 'quasipoisson', data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')))

#w lags
#regsubsets
mod_select <- regsubsets(cnt_HI ~ SST_44 + SST_44_anom + Wind_39 + Upwelling_39 + Upwelling_45 + CBA_South + Wind_44_lag1 + Upwelling_39_lag1 + MEI_lag2 + SST_44_lag2 + Wind_39_lag2 + Upwelling_39_lag2 + Upwelling_45_lag2 + SST_39_anom_lag2 + MEI_lag3 + Wind_44_lag3 + Upwelling_45_lag3, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion'), nbest = 3, really.big = T)
summary(mod_select)

# nb <- summary(glm.nb(cnt_HI ~ SST_44 + SST_44_anom + Wind_39 + Upwelling_39 + Upwelling_45 + CBA_South + Wind_44_lag1 + Upwelling_39_lag1 + MEI_lag2 + SST_44_lag2 + Wind_39_lag2 + Upwelling_39_lag2 + Upwelling_45_lag2 + SST_39_anom_lag2 + Wind_44_lag3, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')))

qp <- summary(glm(cnt_HI ~ Wind_39 + Upwelling_39 + CBA_South + Upwelling_39_lag1 + Wind_39_lag2, data = monthly_anom %>% filter(Pinniped.Common.Name == 'Steller sea lion')), family = 'quasipoisson')

pchisq(deviance(nb), df.residual(nb), lower.tail = F) 
pchisq(deviance(qp), df.residual(qp), lower.tail = F) 

```

2. Are stranding anomalies correlated with certain oceanographic indices?

2a. All cases anomalies
```{r}

# (a) Stranding anoms for combined species 

mod_select <- regsubsets(strand_anom_all ~ NPGO + SST_44_anom + Wind_44 + CBA_South +
          NPGO_lag1 + SST_44_lag1 + SST_39_anom_lag1 + SST_39_lag2 + Wind_39_lag2 + Upwelling_39_lag2 + SST_39_anom_lag2 + Wind_39_lag3 + Period, data = monthly_anom_comb, nbest = 3, really.big = T)
summary(mod_select)

nb <- summary(glm.nb(round(strand_anom_all + 10, 0) ~ Wind_44 + CBA_South + SST_39_anom_lag1 +
          SST_44_lag1 + Wind_39_lag2 + Upwelling_39_lag2 + Wind_39_lag3 + Period, data = monthly_anom_comb))
nb <- summary(glm.nb(round(strand_anom_all + 10, 0) ~ SST_44 + CBA_South +
          Wind_39_lag2 + Upwelling_39_lag2 + Period, data = monthly_anom_comb))
qp <- summary(glm(strand_anom_all + 10 ~ SST_44 + CBA_South +
          Wind_39_lag2 + Upwelling_39_lag2 + Period, family = 'quasipoisson', data = monthly_anom_comb))
pchisq(deviance(nb), df.residual(nb), lower.tail = F) #p = 0.28
pchisq(deviance(qp), df.residual(qp), lower.tail = F) 

# (b) Stranding anoms for separate species

#CSL
 

#HS


#SSL

```

2b. HI cases anomalies
```{r}
#Repeat process from above, waiting for advice on overall approach
# (a) HI case anoms for combined species 

# (b) HI case anoms for separate species

```


3. Are there different models that are a better fit for different species (i.e., are species impacted differently by the same ocean conditions?)
```{r}

```

5. Are stranding hotspots different during EN vs LN conditions? Or some other metric?
```{r}
#Use SatScan to explore 
```


